Modal Logics in the Coq Proof Assistant

### 2014-12-10 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Fix labels and references, and improve the text

\usepackage{fancybox} % Must be before "fancyvrb"
\usepackage[utf8]{inputenc}
\usepackage{url,amsmath,amssymb,fancyvrb,color,graphicx}
\usepackage{proof}
\usepackage{xspace}

@@

Modal logics extend the usual formal logic language with modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, although $A \imp \Box A$ is not necessarily a
theorem. Various informal concepts, such as \emph{necessity and
possibility}, \emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, have been formalized with the help of
modal operators. Therefore, modal logics \cite{ModalLogic} are well-
established for a wide variety of application domains, ranging from
philosophy \cite{ToDo} to formal software verification
\cite{ToDo:TLAPlus}.

@@

One major initial concern was whether the embedding could be a
disturbance to the user. Fortunately, by using \Coq's \texttt{Ltac}
tactic language, we were able to define intuitive new tactics that
hide the technical details of the embedding from the user. The
resulting infra-structure for modal reasoning within \Coq (as
described in Section \ref{sec:modal}) provides a user experience where
modalities can be handled transparently and straightforwardly.
Therefore, a user with basic knowledge of modal logics and \Coq's
tactics should be able to use (and extend) our implementation with no
excessive overhead.

@@

\section{The Embedding of Modal Logics in Coq}
\label{sec:modal}

@@

\section{Tactics for Modalities}

@@

@@

To maximally preserve user intuition in interactive modal logic
theorem proving, the embedding via the possible world semantics should
be as transparent as possible to the user. Fortunately, the basic Coq
tactics described above automatically unfold the shallowest modal
definition in the goal.  Therefore, they can be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers. The situation for the new modal
operators, on the other hand, is not as simple, unfortunately.

@@

\s\s\s\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s\s\s\s\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
$\omega$ must be a fresh name for a box. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are created.
\end{center}

@@

\section{Two Simple Modal Lemmas}

In order to illustrate the tactics described above, 
we show \Coq proofs for two simple but useful modal lemmas. 
The first lemma is a kind of modus ponens with formulas 
under the scope of modal operators.

>>>>

\usepackage{fancybox} % Must be before "fancyvrb"
\usepackage[utf8]{inputenc}
\usepackage{url,amsmath,amssymb,fancyvrb,color,graphicx}
\usepackage{proof}
\usepackage{xspace}
\usepackage{float}

@@

Modal logics extend usual formal logic languages by adding modal
operators ($\Box$ and $\Diamond$) and are characterized by the
\emph{necessitation rule}, according to which $\Box A$ is a theorem if
$A$ is a theorem, although $A \imp \Box A$ is not necessarily a
theorem. Various informal concepts, such as \emph{necessity and
possibility}, \emph{knowledge and belief}, \emph and \emph{temporal
globality and eventuality}, have been formalized with the help of
modal operators. Therefore, modal logics \cite{ModalLogic} are well-
established for a wide variety of application domains, ranging from
philosophy \cite{ToDo} to formal software verification
\cite{ToDo:TLAPlus}.

@@

One major initial concern was whether the embedding could be a
disturbance to the user. Fortunately, by using \Coq's \texttt{Ltac}
tactic language, we were able to define intuitive new tactics that
hide the technical details of the embedding from the user. The
resulting infra-structure for modal reasoning within \Coq (as
described in Section \ref{sec:Embedding}) provides a user experience where
modalities can be handled transparently and straightforwardly.
Therefore, a user with basic knowledge of modal logics and \Coq's
tactics should be able to use (and extend) our implementation with no
excessive overhead.

@@

\section{The Embedding of Modal Logics in Coq}
\label{sec:Embedding}

@@

\section{Tactics for Modalities}
\label{sec:Tactics}

@@

\begin{center}
$\neg A$ is an abbreviation for $A \imp \bot$.
\end{center}

@@

\noindent
To maximally preserve user intuition in interactive modal logic
theorem proving, the embedding via the possible world semantics should
be as transparent as possible to the user. Fortunately, the basic Coq
tactics described above automatically unfold the shallowest modal
definition in the goal.  Therefore, they can be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers. The situation for the new modal
operators, on the other hand, is not as simple, unfortunately.

@@

\s\s\s\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s\s\s\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

@@

\begin{center}
\textbf{eigen-box condition:}\\ 
$\nec_I$ and $\pos_E$ are \emph{strong} modal rules: \\
$\omega$ must be a fresh name for the box they access \\ 
(in analogy to the eigen-variable condition for strong quantifier rules). \\
Every box must be accessed by \emph{exactly one} strong modal inference. \\
\vspace{0.5em}
\textbf{boxed assumption condition:} \\
assumptions should be discharged within the box where they are created.
\end{center}

@@

\section{Two Simple Modal Lemmas}
\label{sec:SimpleExamples}

In order to illustrate the tactics described above, 
we show \Coq proofs for two simple but useful modal lemmas. 
The first lemma resembles modus ponens, but with formulas 
under the scope of modal operators.
